
 

\chapter{Programming and computers through documentaries and movies} 

    
      
 \section{History of informatics}
      1812. Charles Babbage. Differential machine. 
      1872. Ada Lovelace. Daughter of The poet Lord Byron. She invented the first language eith perforated cards. 
      1889. Herman Hollerith. Automatization census. Then 1924, IBM.
      1907 vacuum valve 
      1936. Universal Turing machine. 
      1939. Bombe machine. To decipher enigma. 
      1946. ENIAC. Electronic Numerical Integrator and Computer. Ballistic calculations. 
      1947. Transistor. First assembly language with von Neumann. 
      1960 NASA. Fortran. 
      1977. John Backus. Alan Turing award. Can machines be liberated from the von Neumann style ? 
       
 
  \section{Algorithms and machines: "The imitation game"} 
  
  3 manchester, he was arrested for homosexuality. Homosexuality was punished at that time in england. 
  
  In 1936, Turing had invented a hypothetical computing device that came to be known as the ‘universal Turing machine’. 
  
  Whilst working for the National Physical Laboratory (NPL), Turing published a design for the ACE (Automatic Computing Engine), which was 
  arguably the forerunner to the modern computer. The ACE project was not taken forward, however, and he later left the NPL.
  
   World War II, also known as the Second World War, was a global war that lasted from 1939 to 1945.
  
  1942 he developed the bombe machine to decipher. 
  
  
  The first working device to be built was a point-contact transistor invented in 1947 by American physicists John Bardeen and Walter 
  Brattain 
  while working under William Shockley at Bell Labs. The three shared the 1956 Nobel Prize in Physics for their achievement.
  
  ENIAC Introduced to the world on Feb. 14, 1946, the ENIAC -- Electronic Numerical Integrator and Computer -- was developed by the 
  University 
  of Pennsylvania's John Mauchly and J. Presper Eckert 
  
  was developed under a 1943 contract with the U.S. Army to speed ballistics calculations. It was built under total secrecy and completed 
  only 
  after WWII, however. At 30 feet by 60 feet, weighing 30 tons and using 19,000 vacuum tubes, ENIAC was the epitome of a large system.
  
  The Fleming valve, also called the Fleming oscillation valve, was a thermionic valve or vacuum tube invented in 1904 by English physicist 
  John Ambrose Fleming as a detector for early radio receivers used in electromagnetic wireless telegraphy
  
  \section{First numerical computing: "Hidden figures"}
 
 
  \section{First home computers: "The triumph of nerds"}
  ROBERT CRINKGELY
  
  2.1) Triumph of nerds. Paul allen, bill gates, wozniak, Steve jobs. 
  Mainframe versus personal computer. 
  
  Cobol, fortran, basic. First languages. 
  Nerds thought it was a bible for them. For a 10 years old. It was incredible thrilling experience. 
  
  Vaccum tubes or valves. Microprocessor That's the secret of computer. That's why it's called silicon valley. Intel invented the 
  Microprocessor. 
  They didn't realize the potential of the personal computer
  
  
  
  
  1975 Ed Roberts. Invented the first personal computer ALTAIR 8080 Microprocessor. He invented it to play with. He run a calculator in 
  Albuquerque. No monitor, no keyboard. Homebrew computer club. Stanford. 
  
  To understand what to do with this usesless computer. 
  
  Paul Allan began with the Altair and with a tape of paper programming basic. They were 19. They attached keyboards and monitor. At the end 
  of 1975. A revolution. Groovy
  
  The first apple computer was to impress the club. 
  Two years past from 1975 to build apple 2. Wozniak was the Mozart of the digital computer. 
  Success with 21 years old. 
  
  Visual calc. The first spread sheets.  Profits and expenses. Running numbers by hand. Dan Bricklin. The programmer. They don't make money. 
  They wanted to make a better world. 
  
  Obses with numbers. It was crystal ball. 
  
  Part II
  In 1980 apple ii was a success. An IBM appeared into the scene. At the beginning IBM was making mainframes. Tom Watson. Strict dress code. 
  Work 9 to 5 and Saturday wash the car. Entire culture. The songs of IBM. Gay meant something different. Are you an ibmer. Conservative. IBM 
  noticed the explosion of apple II. 
  
  A pc needs software. Computer language. Operating system. Bill gates has the basic language. CPM Gary kildall. First OS. He was not a 
  fighter. Bill gates was very competitive. 
  IBM was looking for OS and language. Gary didn't receive IBM and bill gates saw the opportunity. They replicated CPM in four months: they 
  call it pc dos 1.0 . They bought it the OS from Tim Patterson. 50 thousand dollars was the price. Then, Bill and Paul become multi 
  billionaire.
  
  IBM entered in the market. 1981. Lotus 1/2/3.
  It was the Killer application. Euphoric. 
  
  People wanted to copy the IBM. 
  Reverse engineering. 
  
  Engineering from Texas computer founded Compaq computer. Micro was availablefrom Intel. Rom bios was patented. You seen the product. You 
  need vigin. Never seen it. After the virgins test. 15 senior engineers. It took one year to accomplish that pc and one million dollars. 
  Terror was in IBM because prices went down because competence. 
  
  IBM with OS/2. Tried to steal the business to microsoft. A culture clash between IBM and microsoft. 1990 microsoft with windows broke 
  relations with IBM but the idea came from apple. 
  
  IBM gave 1/3 to Intell and 1/3 to Microsoft. Big error. 
  
  Part III.
  A revolution to mage pc friendier s
  
  1995 windows announced its OS. Those ideas was invented 20 years ago.  GUI: graphical user interface. 
  
  It all began in 1971. 
  1971 Xerox in Palo Alto. They created the first GUI. Freedom is that company. To work on five years program. Their dreams. The mouse was 
  created there. A terrible mismacth between researchers and managers. 
  
  Steve Jobs wanted to change the world not to earn money. 1979 visit to Xerox. 
  Steve Jobs had the vision . Xerox showed steve three things: 1) a network like internet, 2) a GUI and  3) object oriented programming. He 
  was blinded by 2). It was a turning point. 
  
  After an hour Steve understood what was going on. Apple developed Lisa with one hundred engineers. 
  
  In 1981 apple was in trouble because IBM was selling software that didn't run on MAC. IBM had software. In 1983 they announced a game to 
  create software for mac. Bill Gates was there. He didn't realize that he was going to be his rival Instead of IBM. Machintosh the first 
  affordable pc with a GUI. Very bad sales. It was 1000 dollars more expensive them IBM. 
  
  Commercial against IBM tirany. 
  
  Jobs made agreement with bill gates to create software. He thought that the big blue was the enemy. 
  
  
  What you see is what you get. 
  Wysiwyg. Apple needed a killer application. 
  The problem was the dot printer. Agreement with adobe people to create laser images printings. 
  Adobe also developed in Xerox 
  Apple bought those software.they developed a killing application. Steve terrified everyone. So one steve left apple was twofold. 
  
  Scotty talked with apple board and decided to follow Scotty plans. 
  Steve said that he hired the wrong guy: Scotty Pepsi cola.  Steve left the company. 
  
  1984. Microsoft launched windows to challenge macintosh.  
  
  Apple sewed microsoft. A long a legal battle (6 years). 
   Apple lost. 
  Steve said microsoft have no taste. 
  In the sense that they don't think original ideas. 
  
  Society needs 30 years to arrange Advances in technologies. 
  
  
  
   
   \section{Home computers in Europe: "Clive Sinclair"} 
   
   Micro men.
   He asked for national founds
   
   We exist to push barriers. 
   
   As invertors we are obliged to dream. 
   
   He created sinclair company close to Cambridge university. 
   Clive wanted to develop his electric car and his television but Christopher Curry wanted to create a home computer. 
   
   Margaret thatcher ( Zacher) was elected. 
   
   Curry Christopher began a company with Hermann Hauser an Austrian guy who obtained his ph d in cavendish lab..  Cambridge processor group. 
   They join those guys who did everything for fun. 
   
   Radionics 1978. 
   
   Clive saw the announcement of 2000 pounds of apple ii. And thought. What is it so expensive ? 
   99 ponds. Sinclair zx80. 
   
   His vision a computing device in every home in Britain. Price is the key. 
   
   1980. Sinclair launched zx80 at 99 pounds. 
   
   Acorn was founded by Christopher with memory and more power. 
   
   They began to fight. Thy met in a bar to discuss B BBC computer conquest. Join forces 
   
   Nerds eating with the tester wires.
   
   
   BBC announces a computer home initiative 
   They have a computer series. 
   
   BBC was expecting something different. Chris said that he would do something in four days. 
   He asked his team to do it. They did it. Acorn computer prototype. 
   
   To put a computer in every school of the country. 
   Acorn launched BBC computer. 
   
   125 pounds zx spectrum. 
   It was a revolution. 
   Mainly games. 
   
   The Zilog Z80 is a software-compatible extension and enhancement of the Intel 8080 
   
   What else can I do ? 
   BBC micro from acorn did not have so many games. 
   
   Quantum leap. Another powerful computer is 3 months. 
   
   1984 sinclair quantum leap 399 pounds. Ready to ship in 20 days. 
   
   Clive was concerned with his electric car. 
   
   BBC basic language. It was success. 
   
   
   Sinclair clash with Chris in a pub. 
   
   The computer boom became to an end. 
   
   Acorn designed to ARM microchip. The most famous microchip . 
   As of 2018, over 100.000 million copies of the ARM processor have been manufactured, powering much of the world's mobile computing and 
   embedded systems. Population of the world 
   In 2018: 7600 million persons. 
    
   
   Acorn was considered the british apple. 
   
   Sinclair company was sold to amstrad computers. 
   
   
    
   Z80
   Spectrum 1982 
   
   Acorn atom: competence. 
   
   
   
   Quatum leap 
   
   1988 amstrad. 
   
   The computer boom went to and end. 
   
   The future still needs inventing. 
   
   Acorn sold 1.5 million BBC micros 
   250000 unsold computer electrons. 
   
   The company was bought by Olivetti in 1985. 
   The ARM chip designed by Acorn team has gone on to be the most popular michochip in history. 
   
   Clive Sinclair sold more than 12000 electric mobile cars C5. 
   
   The home computer market is now dominated by giant American companies. 
   
   
  
   
    \section{The birth of internet: "Lo and behold"} 
    Reveries of the connected world.  
    Director: Herzog, Werner. 
    
    
    
    Ucla, 1969 internet was born. 
    First node for decades. 
    Stanford first communication. 
    To log in: connected by a telephone. Lo ang g crashed. 
    
    Arpanet. 197--. Protocol and technology. 
    
    II. The glory of the net. Bio and automobiles. 
    
    III. The dark side. 
    Email with decapitated girl in an car accident. 
    A manifestation of evil. 
    
    IV. Life without the net.  Telescope. 
     Restart. Avoid addiction. 
    In south korea they use diapers no to loose points in the game. 
    
    V. The end of the net. Big flares from the sun. 
    
   
    
   
   
  
  \section{Free software: "The revolution OS. The code"}
  Free software: Richard Stallman 
  
  Open source: ERIC Raymond, Linus Torvald 
  
  
  Bruce Perens: open source definition 
  Coloborating with software. 
  
  Donald Knuth 1979. Latex 
  R Stallman 1983 GNU free software. 
  Eric Steven Raymond : The cathedral and the bazar: musing on Linux and open source by accidental revolutionary. 
  
  Linus Torvalds says that R Stallman was the great philosopher and he was the engineer. 
  
  
  
  
  
  Linux versus windows
  
  
  Free software demands freedom. 
  Open source demands quality. 
  
  People uses programs instead of OS. 
  
  Richard Stallman 
  1971 in MIT. Founded 
  They didn't like passwords. 
  The security was a joke. 
  Homebrew computer club. 
  Bill Gates defends private software. 
  
  Stallman  began to write gnu Unix. 
  
  He finished in 1991. 
  Compilers, editors. 
  With free software you have freedom 
  
  Free has two meanings in English. One 
  Related to money and other related to freedom. 
  
  
  The first company around gnu software wax to sell services around gnu software. In a propietary software, support is a monopoly. Microsoft. 
  
  
  Stallman gives in his manifesto the keys to make money. 
  
  1) support by everyone. Deliver support. 
  cysgnus was the first company. 
  
  1991-1993 infancy of Linux. 
  
  Stallman tried to divide the kernel. Very difficult to debug but very powerful. 
  
  Linus tried a monolithic kernel. 
  
  
  
  The cathedral of the bazaar. He discuss how is possible to develop software by breaking as ll strict software rules of engineers.  
  
  
  They changed free software by open software .
  Stallman is free software and others open software. Free to redistribution. 
  
  Linux open software.
  
  VA Linux company 
  
  IPO 
  
  People wanted to have a Unix sun sparc station at home. That's why Linux was so important.
  It was 2 x times and 1/4 the price of a sun sparc. 
  2000 dollars a IBM PC with Linux in comparison a sun sparc of 7000 dollars. 
  
  Linux had a track similar to internet. In 1993 internet became a common commodity. 
  Apache project. HTTP project. Web server. 
  
  Red hat sotware was one the most important companies to support Linux. 
  
  Linux Torvald created the free kernel. 
  
  
  
  Gnu / Linux was Stallman proposal. 
  
  Apache web server was the Killer application of Linux. It motivates companies to internet. 
  
  Apache software foundations. 
  
  Raymond thought that 
  
  
  Sharing is the basic of society. 
  
  Open software is not comunishm because it does not force people to share. 
  
  The revolution goes prime time. 
  
  Red hat Linux company became public. 
  
  
  GPL Genral public license. 
  
  
  The mind behind Linux 
  
  Linux code. 
  
  
  1998 Netscape was one the first companies.
  Influenedt by the cathedral and the bazaar. 
  Source code was its product. Mozilla product. 
  
  They change to open source 
  1) free redistribution 
  2) 
  3) 
  4) integrity of author code 
  5) no discrimination  
  
  
  Red hat goes to market. 1999
  
  The code: the story of Linux. You tube . Boring  socialism in action. 
  
  Open source project is compared to science or knowledge. 
  
  
  Social machine around open source code 
  
  Revolution OS . You tube. Free software gnu,Linux . Better than the code but boring. 
  
  1991 first Linux 10000 lines of code. 
  
  1992. 40000 lines of code and 1000 users. 
  
  1993. 100000 lines of code and 20000 users. 
  
  1995. 250000 lines of code and 500000 users.
  
  1997. 800000 lines of code and 3.5 million users.  
  
  1998. 1.5 million lines of code and 7.5 million users. 
  
  1999. 12 million users. 
  
  
  
  Gnu : main component c compiler. 
  
  Richard stallman. Founder of gnu. 
  
  Apache web server. 1993. Mass market. 
  Application built with Linux. 
  
  
  The cathedral and the bazzar. Book. Eric Raymond. 
  
  
  
  1998 Netscape. Rel ase source code. Mozzila. Browser. 
  
  1999. 
  
  Influential 
  
  
  VA Linux 
  
  Free software. Gnu. Collaboration. 
  
  Open source. VA Linux. 
  Definition: free redistribution, source code available, derived works permitted, integrity of the author of code. No discrimination against 
  
  Cygnus . Free sotware model.  First company. 
  
  Oracle. 
  
  1998. Venture capilist. Enter in open source. 
  
  When Stallman recieves the prize in that congress. Torvald ignores him. 
  
  VA Linux company multiplied its share values by 10 In one day. 
  
 
 
 
    
  \section{Artificial intelligence: "The thirteenth floor"} 
  
   \section{Math and computers: "The colors of infinity"} 
   
 
  
  \section{Data privacy: "The great hack"}
  
  We are more the commodity. But we are in love with free service. Data attached to our identify. 
  2016 elections. It was a change. 
  
  Project akamo 1million dollars per day in Facebook ads. 
  
  Cambridge analytica. The brain. In UK 
  Alexander Nix was its ceo. 
  
  An environment of great innovation. 
  Expert data science. 
  5000 data points in every american. To model personality. 
  It was tied to the brexit champein. 
  They tried to break the world to rebuild new pieces. 
  Ruining our democracy. 
  
  Brittany Kaiser. Top executive. 
  
  SCL. 
  
  Defeat crooked Hillary : dishonest, illegal. Cambridge analytica created that slogan. 
  
  
  The guardian and the observer. 
  
  It's about our democracy. 
  
  Roger mcnamee. Investor facebook. 
  I feel really guilty. He was not able to sleep at night. 
  
  Brittany says thay obama did not pay her to his champein and she needed money. 
  
  Fear and hate. Divide and conquer. Facebook was created to connect people. 
  
  
  
   Mi
  
  
  
  Carole Cadwalladr. TED talk. She was one of the jurnalist. Driving us apart. Is this what you want. 
  
  You have to understand how you data is affecting your life. Our dignity is at stake. 
  
  Donald Trump 5.9 million versus 66000 dollars Hillary in Facebook ads. 
  
  David Carroll continues to teach and advocate for data rights to be recognized as the new human rights. 
  
  
 
       
 \section{Social networks: "The social dilemma"}
 They are very concerned. 
 When you look around you is like the world is going crazy. 
 
 In 2006 facebook was looking at Google with admiration. They created something good and the same time a machine to make money. 
 
 Book: Jaron Lanier. Ten arguments for deleting your social media accounts. Right now. 
 
 Robert macnameee: selling their users. 
 
 If you  product is free, you are the product. 
 
 What do they do with our data ? 
 They build models that predict our actions. Hi
 
 Facebook techniques: engagement, growth adds
 
 And then give you back the dopamine hit. 
 
 Social media is a drug. That directly affects the release of dopamine in the reward pathway. 
 
 Around 3 hours on average. Social media. 
 
 A pacifier. Social media. 
 
 Hundreds of algorithms running simultaneously 
 
 Few people working on those algorithms. But machine learning has nothing to do with precise  instructions. They don't know the programmers 
 what is going to happen . Because the algorithm is designed to maximize the profit. AI makes the rest. It learns online. And modifies the 
 way of interactions with human beings. In that sense, we don't have the control. 
 
 You are playing with AI. It knows everything about you and you don't know anything about him.  It's not a fair fight. 
 
 Two time scenarios. 
 
 1) when AI will be smarter than human
 
 2) right now . The root of addiction.
  Polarization, radicalization, outrage-ification, vanity-ification. 
 This is checkmate on humanity. 
 
 Example. Imagine wikipedia saying different definitions for different people. If wikipedia could know your profile and it was paid to change 
 a little bit your idea, it would change your information to conquer its objectives. 
 
 Each person has its own reality with its own story based on his profile. 
 You think what they don't realize what is going on. Your question. Because they are not seeing the same information than others. 
 
 Fake news makes companies to earn more money than true news. 
 
 Social media amplifies gossips and rumors not allowing us s to distinguish between true and fake. 
 
 Conspiracy theories. Very easy. 
 
 Technology is the existencial threat. 
 
 Whether it is to be utopia or oblivion will be a touch-and-go relay race right up to the final moment. Humanity is in his final exam as to 
 whether or not it qualifies for continuance in universe. 
 Buckminster Fuller. 
 
 
 Technology is Simultaneously utopia and distopia. 
 
 The business model has a problem. 
 
 These markets undermine democracy and undermine freedom. 
 
 Based on religion: profits at all costs. Corrosive business model.  
 
 Not everyone recognize that we have a problem. 
 
 Reform facebook and Google. 
 
 Qwant in instead of google. Block notifications. 
 
 
 Three basic rules:
 
 1) out of your bedroom
 2) no social media until high school. 
 3) limit you social media time. 
 
 
 Delete your social media accounts. To have more people to have a conversation. 
 
 
 Check mate on humanity 
 
 1) drug addict to social media 
 2) when superintelligence will surpass human beings. 
 3) these markets undermine democracy. 
 
 Bunkminster Fuller 
 
 
 Three rules to convact social media :
 1) avoid phone in your bedroom
 2) don't allow kids to enter in social media until they are at least in high school. 
 3) discuss the time you devote to social media. 
 
 
 
 
 
 